fairmetrics: An R package for group fairness evaluation

### 2025-09-15 utils/joss_paper/paper.tex
recommitting the markdown text

ML models are increasingly used in high-stakes domains such as criminal
justice, healthcare, finance, employment, and education (Mehrabi et al.
2021; Mattu 2016; Gao et al. 2024). Existing fairness evaluation
software report point estimates and/or visualizations, without any
measures of uncertainty. This limits users' ability to determine whether
observed disparities are statistically significant. \texttt{fairmetrics}
addresses this limitation by including confidence intervals for both
difference and ratio based fairness metrics to enable more robust and
statistically grounded fairness assessments.

@@

While the choice of fairness metric used is context dependent, we show
all criteria available with the \texttt{get\_fairness\_metrics()}
function for illustrative purposes. In this example, we evaluate the
model's fairness with respect to the binary protected attribute
\texttt{gender}. The model is trained on a subset of the data and the
predictions are made and evaluated on a test set. A statistically
significant difference across groups at a given level of significance is
indicated when the confidence interval for a difference-based metric
does not include zero or when the interval for a ratio-based metric does
not include one.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Train a classification model (e.g., random forest).}
\CommentTok{\# Add the vector of predicted probabilities to the test data}
\CommentTok{\# to evaluate fairness.}
\FunctionTok{library}\NormalTok{(fairmetrics)}
\CommentTok{\# Setting alpha=0.05 for 95\% confidence intervals}
\FunctionTok{get\_fairness\_metrics}\NormalTok{(}
 \AttributeTok{data =}\NormalTok{ test\_data,}
 \AttributeTok{outcome =} \StringTok{"day\_28\_flg"}\NormalTok{,}
 \AttributeTok{group =} \StringTok{"gender"}\NormalTok{,}
 \AttributeTok{probs =} \StringTok{"pred"}\NormalTok{,}
 \AttributeTok{cutoff =} \FloatTok{0.41}\NormalTok{, }
 \AttributeTok{alpha =} \FloatTok{0.05}
\NormalTok{)}

@@

Other R packages similar to \texttt{fairmetrics} include
\texttt{fairness} (Kozodoi and V. Varga 2021), \texttt{fairmodels}
(Wiśniewski and Biecek 2022) and \texttt{mlr3fairness} (Pfisterer, Siyi,
and Lang 2024). \texttt{fairmetrics} differs from these packages in two
ways. The first difference is that \texttt{fairmetrics} calculates ratio
and difference-based group fairness metrics and their corresponding
confidence intervals, allowing for more meaningful inferences about the
fairness criteria. The second difference is that \texttt{fairmetrics}
does not possess any external dependencies and has a lower memory
footprint.
\hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows
the comparison of memory used and dependencies required when loading
each library.

>>>>

ML models are increasingly used in high-stakes domains such as criminal
justice, healthcare, finance, employment, and education (Mehrabi et al.
2021; Mattu 2016; Gao et al. 2024). Existing fairness evaluation
software report point estimates and/or visualizations, without any
measures of uncertainty. This limits users' ability to determine whether
observed disparities are statistically significant. \texttt{fairmetrics}
addresses this limitation by including confidence intervals (CIs) for
both difference and ratio based fairness metrics to enable more robust
and statistically grounded fairness assessments.

@@

While the choice of fairness metric used is context dependent, we show
all criteria available with the \texttt{get\_fairness\_metrics()}
function for illustrative purposes. In this example, we evaluate the
model's fairness with respect to the binary protected attribute
\texttt{gender}. The model is trained on a subset of the data and the
predictions are made and evaluated on a test set. A statistically
significant difference across groups at a given level of significance is
indicated when the CI for a difference-based metric does not include
zero or when the interval for a ratio-based metric does not include one.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Train a classification model (e.g., random forest).}
\CommentTok{\# Add the vector of predicted probabilities to the test data}
\CommentTok{\# to evaluate fairness.}
\FunctionTok{library}\NormalTok{(fairmetrics)}
\CommentTok{\# Setting alpha=0.05 for 95\% CIs}
\FunctionTok{get\_fairness\_metrics}\NormalTok{(}
 \AttributeTok{data =}\NormalTok{ test\_data,}
 \AttributeTok{outcome =} \StringTok{"day\_28\_flg"}\NormalTok{,}
 \AttributeTok{group =} \StringTok{"gender"}\NormalTok{,}
 \AttributeTok{probs =} \StringTok{"pred"}\NormalTok{,}
 \AttributeTok{cutoff =} \FloatTok{0.41}\NormalTok{, }
 \AttributeTok{alpha =} \FloatTok{0.05}
\NormalTok{)}

@@

Other R packages similar to \texttt{fairmetrics} include
\texttt{fairness} (Kozodoi and V. Varga 2021), \texttt{fairmodels}
(Wiśniewski and Biecek 2022) and \texttt{mlr3fairness} (Pfisterer, Siyi,
and Lang 2024). \texttt{fairmetrics} differs from these packages in two
ways. The first difference is that \texttt{fairmetrics} calculates ratio
and difference-based group fairness metrics and their corresponding CIs,
allowing for more meaningful inferences about the fairness criteria. The
second difference is that \texttt{fairmetrics} does not possess any
external dependencies and has a lower memory footprint.
\hyperref[tab:memory_dep_usage]{Table~\ref*{tab:memory_dep_usage}} shows
the comparison of memory used and dependencies required when loading
each library.
